### 2020-06-07 report/sections/server/react/efsm.tex
Report: progress on chapter 5 runtime
Report: complete chapter 5

\begin{lstlisting}[language=javascript,numbers=none]
type MessageHandler = (message: any) => void;
\end{lstlisting}

@@

We generate the \texttt{handle()} method for each
abstract receive state component for state-specific message
parsing and processing. We show the full generated code
for state S42 in \cref{lst:reactreceivestate}.
Unlike \fancyname{NodeMPST},
the \texttt{handle()} method (\cref{line:reactreceivehandle})
returns the successor state identifier to the runtime
to make the transition.
We leverage the discriminated union from our
generated message types (\cref{lst:reactreceivetypes})
to invoke the correct receive handler (\cref{line:reactreceivediscriminate}).

@@

	handle(message: any): State { (*@\label{line:reactreceivehandle}@*)
		const parsedMessage = JSON.parse(message) as Message;
		switch (parsedMessage.label) { (*@\label{line:reactreceivediscriminate}@*)
			case Labels.THANKS:
				this.THANKS(...parsedMessage.payload);
				return TerminalState.S41;
			case Labels.TERMINATE:
				this.TERMINATE(...parsedMessage.payload);
				return TerminalState.S41;
		}
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Code for \trole{Client} 
Receive State S42 in \tprotocol{Adder} protocol}
\label{lst:reactreceivestate}
\end{figure}

>>>>

\begin{lstlisting}[language=javascript,numbers=none]
type MessageHandler = (message: any) => State;
\end{lstlisting}

@@

We generate the \texttt{handle()} method for each
abstract receive state component for state-specific message
parsing and processing. We show the full generated code
for state S42 in \cref{lst:reactreceivecomponent}.
Unlike \fancyname{NodeMPST},
the \texttt{handle()} method (\cref{line:reactreceivehandle})
returns the successor state identifier to the runtime
to make the transition.
We leverage the discriminated union from our
generated message types (\cref{lst:reactreceivetypes})
to invoke the correct receive handler (\cref{line:reactreceivediscriminate}).

@@

	handle(message: any): State { (*@\label{line:reactreceivehandle}@*)
		const parsedMessage = JSON.parse(message) as Message;
		switch (parsedMessage.label) { (*@\label{line:reactreceivediscriminate}@*)
			case Labels.THANKS:
				this.THANKS(...parsedMessage.payload);
				return TerminalState.S41;
			case Labels.TERMINATE:
				this.TERMINATE(...parsedMessage.payload);
				return TerminalState.S41;
		}
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Code for \trole{Client} 
Receive State S42 in \tprotocol{Adder} protocol}
\label{lst:reactreceivecomponent}
\end{figure}

### 2020-06-07 report/sections/server/react/limitations.tex
Report: complete chapter 5

\section{Limitations}
As discussed, our EFSM encoding provides \textbf{affine} channel
usage guarantees for send states as supposed to \textit{linear} usage
guarantees. Suppose we bind a send action to a
button click event -- strictly speaking,
even if we guarantee that the button can be clicked exactly once,
we cannot guarantee that the button will ever be clicked by the user.
However, existing work \cite{PureScript2019,MVU2020} 
share the same limitation, and we view this as an
inherent limitation in front-end development, since there is
a limit to the types of assumptions we can make about how the
user interacts with the browser.
Redefining the definition of channel linearity 
for GUI programming use cases
would be an interesting discussion.

Our runtime implementation also limits the developer from
passing data into their EFSM state component through props.
This is because the runtime takes full control of which
state component to render, and by definition, what props 
are passed to the EFSM state components.
However, we equally view this as a way to encourage developers
to decouple communication state management from
business logic state management, in the spirit
of separating concerns.
In \cref{chap:eval},
we present examples of developer implementations
that use separate abstractions (such as the React Context API or
the third-party Redux library) for propagating 
stateful application data in \cref{chap:eval}.

>>>>

\section{Limitations}
Our EFSM encoding provides \textbf{affine} channel
usage guarantees for send states as supposed to \textit{linear} usage
guarantees. Suppose we bind a send action to a
button click event -- strictly speaking,
even if we guarantee that the button can be clicked exactly once,
we cannot guarantee that the button will ever be clicked by the user.
However, existing work \cite{PureScript2019,MVU2020} 
share the same limitation, and we view this as an
inherent limitation in front-end development, since there is
a limit to the types of assumptions we can make about how the
user interacts with the browser.
Redefining the definition of channel linearity 
for GUI programming use cases
would be an interesting discussion.

Our runtime implementation also limits the developer from
passing data into their EFSM state component through props.
This is because the runtime takes full control of which
state component to render, and by definition, what props 
are passed to the EFSM state components.
However, we equally view this as a way to encourage developers
to decouple communication state management from
business logic state management, in the spirit
of separating concerns.
In \cref{chap:eval},
we present examples of developer implementations
that use separate abstractions (e.g. the \textit{React Context API} or
the third-party \textit{Redux} state management framework) for propagating 
stateful application data in \cref{chap:eval}.

### 2020-06-07 report/sections/server/react/runtime.tex
Report: complete chapter 5

We elaborate on \dots

\subsection{Sending Messages}
\begin{itemize}
\item passes in the factory prop, which is a factoryfactory to be precise;
\item sendMessage is bound in the closure and sends the message in the same way as how Node codegen does, by convention
\item (if not already mentioned in the node section) send is asynchronous
\end{itemize}

\begin{lstlisting}[language=javascript,tabsize=2]
private buildFactory<T>(label: string, successor: State) {
	return <K extends keyof DOMEvents>
				 (ev: K, handler: EventHandler<E, K> => {
		const send = (payload: T) => 
			this.sendMessage(label, payload, successor);
		return class extends React.Component {
			render() {
				const props = {
					[ev as string]: (e: FunctionArguments<DOMEvents[K]>) => {
						const payload = handler(e);
						send(payload);
					}
				};
				return <div {...props}>{this.props.children}</div>
			}		
		}	
	}
}
\end{lstlisting}

\subsection{Receiving Messages}
\begin{itemize}
\item same edge case as the back-end code generation, as message can arrive before handlers are registered
\item use the same two-queue mechanism to keep track of `what comes first' -- i.e. if message comes first, then put in message queue, so handler can pop off when registered; vice-versa, if handler is registered first, put in handler queue so message can be processed when that is received
\item importantly -- only one queue is required, as server-centric topologies only have the client talk to the server
\end{itemize}

\subsection{Handling Termination}
\begin{itemize}
\item close the websocket with a clean exit code
\end{itemize}

>>>>

We keep track of the current EFSM state in the state of the
private \texttt{Client} component.
This enables the UI to be re-rendered, and therefore ``react'',
to EFSM state transitions, which is precisely the goal of \fancyname{ReactMPST}.

\subsection{Sending Messages}
Messages are sent through the WebSocket in the same way 
as \fancyname{NodeMPST} -- objects marked with label and payload
are serialised into JSON. The process of sending is always followed
by a transition to the successor state, so we define the
\texttt{sendMessage()} method on the runtime to also take
the successor state as a function argument, so it can make that transition
immediately after.

\begin{lstlisting}[language=javascript,numbers=none]
private sendMessage(label: string, payload: any, next: State) {
	this.props.ws.send(JSON.stringify({ label, payload }));
	this.advance(next);
}
\end{lstlisting}

We focus on the higher-order factory that the runtime passes
to the send component as a prop. The implementation shown in
\cref{lst:reactsendfactory} is less straightforward,
so we will walk through the steps.

The prop allows the send component to build factory functions
for sending messages with a particular \texttt{label}
followed by a transition to some \texttt{successor} state
(\cref{line:sendfactoryparams}).
The higher-order factory relies on \textit{closures}
to bind the channel action into the generated component.
Here, \texttt{buildFactory} returns a (factory) function that
builds an anonymous React component which wraps the 
component's children into a \texttt{<div>} with special props 
(\cref{line:sendfactorydiv}).
The props for this anonymous React component comprises of a new
event listener bound to the UI event (\texttt{ev})
specified by the developer (such as \lstonelinejs{'onClick'}).
This new event listener is defined in terms of the
\texttt{handler} from the developer's implementation:
\textbf{(1)} it invokes the handler to obtain the payload to send 
(\cref{line:sendfactoryhandler}), which is constrained by
the generic type parameter to ensure it matches the protocol specification;
then, \textbf{(2)} it sends the message, which triggers the EFSM
to advance to the successor state.
We cannot directly call \lstonelinejs{this.sendMessage} inside
the generated React component because \lstonelinejs{this} refers
to the anonymous class; instead, we define a partially applied
version of the \texttt{sendMessage} method as a local variable 
(\cref{line:sendfactorycurrysend}).

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
private buildFactory<T>(label: string, successor: State) { (*@\label{line:sendfactoryparams}@*)
	return <K extends keyof DOMEvents>
				 (ev: K, handler: EventHandler<E, K> => {
		const send = (payload: T) => (*@\label{line:sendfactorycurrysend}@*)
			this.sendMessage(label, payload, successor);
		return class extends React.Component {
			render() {
				const props = {
					[ev as string]: (e: FunctionArguments<DOMEvents[K]>) => {
						const payload = handler(e); (*@\label{line:sendfactoryhandler}@*)
						send(payload);
					}
				};
				return <div {...props}>{this.props.children}</div> (*@\label{line:sendfactorydiv}@*)
			}		
		}	
	}
}
\end{lstlisting}
\captionof{lstlisting}{Higher-Order Factory Function in 
Browser-Side Runtime}
\label{lst:reactsendfactory}
\end{figure}

\subsection{Receiving Messages}

We take the same approach as \fancyname{NodeMPST} --
the runtime keeps track of the ``active'' receive handler,
and the \texttt{onmessage} event listener for the WebSocket
is defined to dynamically process the incoming message
using the active receive handler.
We also need to address the same consistency problem, since
a message arrival event can be handled \textit{before}
the receive handler is registered.
The receive handler is registered when the receive state component
is rendered on the DOM, which occurs when \lstonelinejs{this.setState}
is invoked in the transition function.
\lstonelinejs{this.setState} may be \textit{asynchronous}
as React will internally batch multiple calls into a single update
for performance \cite{ReactState}, so it is possible for the WebSocket message
event to be queued (and hence, procesesd) before the receive handler
is registered.

We address this problem using a pair of queues to keep track
of messages waiting for handlers, and handlers waiting for messages.
Unlike \fancyname{NodeMPST}, we do \textit{not} need a mapping
of queues because browser-side endpoints only receive from the server.

\begin{lstlisting}[language=javascript,numbers=none]
private messageQueue: any[];
private handlerQueue: MessageHandler[];
\end{lstlisting}

This lets us define the WebSocket \texttt{onmessage} event
listener and the \texttt{register} function (which we pass as a prop
to the receive state component) in a similar way compared with
\fancyname{NodeMPST} -- we show this in \cref{lst:reactreceiveruntime}.
Recall that the message handler registered
by the receive state component returns the successor state,
so we invoke the transition function on the result directly 
(\cref{line:reactreceiveadvance1,line:reactreceiveadvance2}).

\begin{figure}[!h]
\begin{lstlisting}[language=javascript]
private register(handler: MessageHandler) {
	const message = this.messageQueue.shift();
	if (message !== undefined) { this.advance(handler(message)); } (*@\label{line:reactreceiveadvance1}@*)
	else { this.handlerQueue.push(handler); }
}

private receive({ data }: MessageEvent) {
	const handler = this.handlerQueue.shift();
	if (handler !== undefined) { this.advance(handler(data); } (*@\label{line:reactreceiveadvance2}@*)
	else { this.messageQueue.push(data); }
}
\end{lstlisting}
\captionof{lstlisting}{Receive Handler Registration and Message Event
Listener in \fancyname{ReactMPST}}
\label{lst:reactreceiveruntime}
\end{figure}

\subsection{Handling Termination}

The terminal state component notifies the runtime (through
the \texttt{terminate} function passed down as a prop)
to close the connection. The implementation is straightforward --
we simply invoke the \texttt{close()} method on the WebSocket.

\begin{lstlisting}[language=javascript,numbers=none]
private terminate() {
	this.props.ws.close();
}
\end{lstlisting}
